---
title: "Getting Started with streamMetabolizer"
author: "Alison Appling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{get_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

# Introduction

This file demonstrates how to estimate stream metabolism using
`streamMetabolizer` and some example data supplied by Bob Hall for French Creek
in Laramie, WY.

# Installation

## Recommended

The most stable+current version of this package can be installed with this R command:
```{r, eval=FALSE}
install.packages("streamMetabolizer", dependencies=TRUE, 
  repos=c("https://owi.usgs.gov/R","https://cran.rstudio.com"))
```
and updated with this command:
```{r, eval=FALSE}
update.packages(oldPkgs=c("streamMetabolizer","unitted"),
  dependencies=TRUE, repos=c("https://owi.usgs.gov/R", "https://cran.rstudio.com"))
```

## For the adventurous

The in-development version of the package can be installed with the `devtools` package. 
We can make no guarantees about the stability of this version, 
but it might have new features that you'll like.
If you go this route, you will need to install the package dependencies separately, like this:
```{r, eval=FALSE}
install.packages(
  c("LakeMetabolizer","unitted","dplyr","lazyeval","lubridate","magrittr",
    "tidyr","chron","dygraphs","ggplot2","RCurl","rstan","XML","xts"),
  repos=c("https://owi.usgs.gov/R","https://cran.rstudio.com"))
```
You can then install the most cutting edge version of streamMetabolizer with this command:
```{r, eval=FALSE}
devtools::install_github("USGS-R/streamMetabolizer", ref="develop")
```

## Software dependencies

If you plan to use Bayesian models, you will need an up-to-date installation of [Rtools](https://cran.r-project.org/bin/windows/Rtools/). Run `devtools::find_rtools()` to make sure Rtools is ready to go. (Rtools is broadly useful for R packages and might become a stronger dependency of `streamMetabolizer` in the future.) Having Rtools installed will allow you to install rstan, the package that `streamMetabolizer` relies on to run MCMC models.


# Preparation

## Load the input data

Load the R libraries that will be used.
```{r libs, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(streamMetabolizer)
})
```

Load a moderately-sized example dataset from the package (data are from French
Creek in Laramie, WY, courtesy of Bob Hall). Explicitly attached units are
optional but useful for confirming that the data are in the expected format.
We'll use the `streamMetabolizer` standard in defining our day to run from 4 am
(`day_start=4`) to 4 am (`day_end=28`).
```{r data}
dat <- data_metab(num_days='3', res='15', day_start=4, day_end=28, attach.units=TRUE)
```

First inspect the dimensions, column names, and units of the data. Note that
`unitted` data, such as that returned by `data_metab(..., attach.units=TRUE)`,
prints out with a second row called 'U' giving the units for each column.
```{r data_check}
dim(dat)
dat[c(1,48,96,240,288),] # some example rows
```

## Explore the input data

Now graphically inspect the input data.
```{r viz_inputs_DO, fig.width=7, fig.height=3}
dat %>% unitted::v() %>%
  mutate(DO.pctsat = 100 * (DO.obs / DO.sat)) %>%
  select(solar.time, starts_with('DO')) %>%
  gather(type, DO.value, starts_with('DO')) %>%
  mutate(units=ifelse(type == 'DO.pctsat', 'DO\n(% sat)', 'DO\n(mg/L)')) %>%
  ggplot(aes(x=solar.time, y=DO.value, color=type)) + geom_line() + 
  facet_grid(units ~ ., scale='free_y') + theme_bw() +
  scale_color_discrete('variable')
```

```{r viz_inputs_other, fig.width=7, fig.height=4}
labels <- c(depth='depth\n(m)', temp.water='water temp\n(deg C)', light='PAR\n(umol m^-2 s^-1)')
dat %>% unitted::v() %>%
  select(solar.time, depth, temp.water, light) %>%
  gather(type, value, depth, temp.water, light) %>%
  mutate(
    type=ordered(type, levels=c('depth','temp.water','light')),
    units=ordered(labels[type], unname(labels))) %>%
  ggplot(aes(x=solar.time, y=value, color=type)) + geom_line() + 
  facet_grid(units ~ ., scale='free_y') + theme_bw() +
  scale_color_discrete('variable')
```

## Check the input data format

Our data need to have specific column names and units to fit an MLE, Bayesian,
or nighttime-regression model. To see what is required, use the `metab_inputs`
function to get a description of the required inputs for a given model type. The
output of metab_inputs is a table describing the required column names, the
classes and units of the values in each column, and whether that column is
required or optional. The inputs are identical for the model types 'mle',
'bayes', and 'night', so here we'll just print the requriements for 'mle'.
```{r data_needs}
metab_inputs('mle', 'data')
```

Our data format matches that of the requirements from `metab_inputs`:
```{r data_str}
data.frame(
  colname=names(dat), 
  class=unname(sapply(unitted::v(dat), function(col) paste(class(col), collapse=','))),
  units=unname(unitted::get_units(dat)))
```

We're ready to go.

# Modeling Overview

There are three steps to fitting a metabolism model in `streamMetabolizer`.

1. Identify the name of the model structure you want using `mm_name()`.
2. Set the specifications for the model using defaults from`specs()` as a starting point.
3. Fit the model with `metab()`.

And then a fourth step once you've fit a model.

4. Inspect the output with functions including `get_params()`, `get_fit()`, 
`predict_metab()`, `predict_DO()`, `plot_metab_preds()`, and `plot_DO_preds()`.

We will walk through these steps twice in this tutorial to demonstrate two
different models applied to the same dataset.

# A Maximum Likelihood Model

## 1. Choose a model structure

A model is fully specified by a concise if cryptic model name. Examples are 
"b_np_oipi_tr_plrckm.stan" and "m_np_oi_tr_plrckm.nlm". You never need to
remember these concise names; instead, you can construct and interpret such
names using the `mm_name` and `mm_parse_name` functions, respectively.

`streamMetabolizer` offers three core modeling `type`s: maximum likelihood
estimation ('mle'), hierarchical or non-hierarchical Bayesian models ('bayes'),
or nighttime regression ('night'). It is always helpful in `mm_name` to specify
the model `type` you want. Further structural details can be specified in
`mm_name` or left to the defaults.

Here are some example calls to create and parse model names:
```{r names}
three_names <- c(
  mm_name(type='mle'), # the default MLE model
  mm_name(type='mle', ode_method='euler'), # override the default ode_method
  mm_name(type='bayes')) # the default Bayesian model
three_names

# parse the above model names
mm_parse_name(three_names)
```

See the help file at `?mm_name` for a description of the arguments to `mm_name`
and results from `mm_parse_name`.

Here we will fit the default MLE model. Many others are available (see Other
Models below), but this one is common and fast. Call `mm_name` again to set the
default MLE model name/structure.
```{r mle_name}
mle_name <- mm_name(type='mle')
mle_name
```

## 2. Set the specifications

Having chosen a model, we next need to define a list of specifications for that
model. The `specs` function creates a list appropriate to the model we chose.
```{r mle_specs}
mle_specs <- specs(mle_name)
mle_specs
```

See `?specs` for definitions of all specifications. Note that most of the 
specifications in that help file are omitted from the output of
`specs(mle_name)` above - this is because MLE models are simple and don't have
many parameters to set. Any of those parameters that are included in
`mle_specs` can be modified, either by calling `specs()` again or by replacing
that value in the `mle_specs` list. Here is a command that sets the the inital
values of GPP, ER, and K600 for the likelihood maximization. (I've done this
just for illustration; the model results aren't affected by these particular
changes for this particular dataset.)
```{r specs_details}
mle_specs <- specs(mle_name, init.GPP.daily=2, init.ER.daily=-1, init.K600.daily=3)
```

## 3. Fit the model

Now actually fit the model using the `metab` function.
```{r mle_fit, warning=FALSE}
mle_fit <- metab(mle_specs, data=dat, info=c(site='French Creek, WY', source='Bob Hall'))
```

It's optional, but sometimes helpful, to include some sort of metadata in the
`info`, as I've done above. I've chosen to put the metadata in a character
vector, but metadata can take any format you like.

## 4. Inspect the output

Models show lots of relevant information if you simply print them at the command line.
```{r show}
mle_fit
```

You can also extract specific pieces of information using designated accessor
functions. For example, the `info` and `data` are saved in the fitted model
object and can be pulled out with `get_info` and `get_data`, respectively.
```{r info}
get_info(mle_fit)
head(get_data(mle_fit))
```

We can also get information about the model fitting process.
```{r info2}
get_fitting_time(mle_fit) # the time it took to fit the model
get_version(mle_fit) # the streamMetabolizer version used to fit the model
get_specs(mle_fit) # the specifications we passed in
```

There is a function to plot the daily metabolism estimates.
```{r plot_metab1, fig.width=7, fig.height=4.5}
plot_metab_preds(mle_fit)
```

There is also a function to plot the dissolved oxygen predictions (lines) along
with the original observations (points).
```{r plot_metab2, fig.width=7, fig.height=6}
plot_DO_preds(mle_fit)
```

You can output the daily and instantaneous predictions to data.frames for
further inspection.
```{r pred_dfs}
met_preds <- predict_metab(mle_fit)
met_preds
oxy_preds <- predict_DO(mle_fit)
head(oxy_preds)
```

# A Bayesian Model

## 1. Choose a model structure

For our second modeling example, we will specify a Bayesian model with both 
observation error and process error. We won't pool K600 here because we don't 
have many days of data, but pooling is one feature that makes Bayesian models 
better than MLE models in general. Another great feature of Bayesian models is 
that they produce more accurate and nuanced confidence intervals.
```{r bayes_name}
bayes_name <- mm_name(
  type='bayes', pool_K600='none', 
  err_obs_iid=TRUE, err_proc_acor=FALSE, err_proc_iid=TRUE, 
  ode_method='trapezoid')
bayes_name
```

## 2. Set the specifications

Just as for the previous model, we now pass the model name to `specs()` to get a
list of default specifications for this model. Note that the length and contents
of this new specifications list differs substantially from the list we had for
the MLE model.
```{r bayes_specs}
bayes_specs <- specs(bayes_name)
bayes_specs
```

At this point we can alter some of the specifications if desired. I'm choosing a
very small number of `burnin_steps` and `saved_steps` because I don't want to 
wait long for the vignette to run. When you run your own models, consider 
bumping those numbers up substantially (to several thousand or so, depending on 
exactly which model you're using.)
```{r bayes_specs2}
# one way to alter specifications: call specs() again
bayes_specs <- specs(bayes_name, burnin_steps=100, saved_steps=200, n_cores=1, GPP_daily_mu=3, GPP_daily_sigma=2)
# another way: use revise()
bayes_specs <- revise(bayes_specs, burnin_steps=100, saved_steps=200, n_cores=1, GPP_daily_mu=3, GPP_daily_sigma=2)
```

## 3. Fit the model

We now fit the model with the new specifications. Bayesian models take longer to
run, so be patient.
```{r bayes_fit}
bayes_fit <- metab(bayes_specs, data=dat)
```

## 4. Inspect the output

All of the same inspection functions are available as before, including
`get_fitting_time`, `get_info`, `predict_metab`, and so on. And of course
you can start by simply printing the model to the console.
```{r}
bayes_fit
```

One slight difference is that hierarchical Bayesian models sometimes have 
general warnings and errors not specific to any one day. If there were any, you 
would see a note in the 'warnings' or 'errors' columns of `preds`, and you could
see the full message[s] as elements in the list returned by `get_fit()`. There's
also a [smaller] possibility of errors or warnings directly associated with 
metabolism prediction rather than fitting, inspectable in the 'warnings' and
'errors' columns of the data.frame returned by `predict_metab()`.
```{r bayes_warning}
# here's where you'd find fitting messages:
select(get_params(bayes_fit), warnings, errors)
get_fit(bayes_fit)$warnings
get_fit(bayes_fit)$errors
# and prediction messages
select(predict_metab(bayes_fit), warnings, errors)
```

Here are the predictions from this last model:
```{r bayes_pred_tbl, results='asis'}
predict_metab(bayes_fit) %>% 
  lapply(function(col) if(is.numeric(col)) round(col, 2) else col ) %>%
  as.data.frame() %>%
  knitr::kable()
```

```{r, fig.width=7, fig.height=6}
plot_DO_preds(bayes_fit)
```

# Other Models

Many model structures are available. We have only demonstrated two above (one
MLE and one Bayesian model). You can see the full list of model options by 
calling `mm_valid_names()` (here we'll just print a sampling of the full list):

```{r}
valid_names <- mm_valid_names(type=c('bayes','mle','night'))
length(valid_names)
c(valid_names[seq(1,length(valid_names),length.out=20)], '...')
```

Remember that any of these structures can be specified by using the model name 
directly or passing the corresponding arguments to `mm_name` to create the name.
See `?mm_name` for a detailed explanation of the options, and/or
`mm_parse_name()` to split a cryptic model name into its component features.

You're on your way!
