---
title: "Getting Started with streamMetabolizer"
author: "Alison Appling"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{getstarted}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

## Introduction

This file demonstrates how to estimate stream metabolism using
`streamMetabolizer` and some example data supplied by Bob Hall for French Creek
in Laramie, WY.

## Install

### Recommended

The most stable+current version of this package can be installed with this R command:
```{r, eval=FALSE}
install.packages("streamMetabolizer", dependencies=TRUE, 
  repos=c("http://owi.usgs.gov/R","https://cran.rstudio.com"))
```
and updated with this command:
```{r, eval=FALSE}
update.packages(oldPkgs=c("streamMetabolizer","unitted"),
  dependencies=TRUE, repos=c("http://owi.usgs.gov/R", "https://cran.rstudio.com"))
```

### For the adventurous

The in-development version of the package can be installed with the `devtools` package. 
We can make no guarantees about the stability of this version, 
but it might have new features that you'll like.
If you go this route, you will need to install the package dependencies separately, like this:
```{r, eval=FALSE}
install.packages(
  c("LakeMetabolizer","unitted","dplyr","lazyeval","lubridate","magrittr",
    "tidyr","chron","dygraphs","ggplot2","RCurl","runjags","rstan","XML","xts"),
  repos=c("http://owi.usgs.gov/R","https://cran.rstudio.com"))
```
You can then install the most cutting edge version of streamMetabolizer with this command:
```{r, eval=FALSE}
devtools::install_github("USGS-R/streamMetabolizer", ref="develop")
```

### Software dependencies

If you plan to use Bayesian models with 
`engine='jags'`, you will also need to install [JAGS](http://mcmc-jags.sourceforge.net/). (JAGS may be deprecated in the future because so far we've seen better performance with Stan. You should be able to switch seamlessly between the two MCMC engines.)

If you plan to use Bayesian models with 
`engine='stan'`, you will need an up-to-date installation of [Rtools](http://cran.r-project.org/bin/windows/Rtools/). Run `devtools::find_rtools()` to make sure Rtools is ready to go. (Rtools is broadly useful for R packages and might become a stronger dependency of `streamMetabolizer` in the future.)


### Load

Load the R libraries that will be used.
```{r libs, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(ggplot2)
  library(streamMetabolizer)
})
```

Load a moderately-sized example dataset from the package (data are from French
Creek in Laramie, WY, courtesy of Bob Hall). Explicitly attached units are
optional but useful for confirming that the data are in the expected format.
We'll use the `streamMetabolizer` standard in defining our day to run from 4 am
(`day_start=4`) to 4 am (`day_end=28`).
```{r data}
dat <- data_metab(num_days='3', res='15', day_start=4, day_end=28, attach.units=TRUE)
```

First inspect the dimensions, column names, and units of the data. Note that
`unitted` data, such as that returned by `data_metab(..., attach.units=TRUE)`,
prints out with a second row called 'U' giving the units for each column.
```{r data_check}
dim(dat)
dat[c(1,48,96,240,288),] # some example rows
```

### Explore the input data

Now graphically inspect the input data.
```{r viz_inputs_DO, fig.width=7, fig.height=3}
dat %>% unitted::v() %>%
  mutate(DO.pctsat = 100 * (DO.obs / DO.sat)) %>%
  select(solar.time, starts_with('DO')) %>%
  gather(type, DO.value, starts_with('DO')) %>%
  mutate(units=ifelse(type == 'DO.pctsat', 'DO\n(% sat)', 'DO\n(mg/L)')) %>%
  ggplot(aes(x=solar.time, y=DO.value, color=type)) + geom_line() + 
  facet_grid(units ~ ., scale='free_y') + theme_bw() +
  scale_color_discrete('variable')
```

```{r viz_inputs_other, fig.width=7, fig.height=4}
labels <- c(depth='depth\n(m)', temp.water='water temp\n(deg C)', light='PAR\n(umol m^-2 s^-1)')
dat %>% unitted::v() %>%
  select(solar.time, depth, temp.water, light) %>%
  gather(type, value, depth, temp.water, light) %>%
  mutate(
    type=ordered(type, levels=c('depth','temp.water','light')),
    units=ordered(labels[type], unname(labels))) %>%
  ggplot(aes(x=solar.time, y=value, color=type)) + geom_line() + 
  facet_grid(units ~ ., scale='free_y') + theme_bw() +
  scale_color_discrete('variable')
```

### Check the input data format

Our data need to have specific column names and units to fit an MLE, Bayesian,
or nighttime-regression model. To see what is required, use the `metab_inputs`
function to get a description of the required inputs for a given model type. The
output of metab_inputs is a table describing the required column names, the
classes and units of the values in each column, and whether that column is
required or optional. The inputs are identical for the model types 'mle',
'bayes', and 'night', so here we'll just print the requriements for 'mle'.
```{r data_needs}
metab_inputs('mle', 'data')
```

Our data format matches that of the requirements from `metab_inputs`:
```{r data_str}
data.frame(
  colname=names(dat), 
  class=unname(sapply(unitted::v(dat), function(col) paste(class(col), collapse=','))),
  units=unname(unitted::get_units(dat)))
```

We're ready to go.

## Modeling Overview

There are three steps to fitting a metabolism model in `streamMetabolizer`.

1. Identify the name of the model structure you want using `mm_name()`.
2. Set the specifications for the model using defaults from`specs()` as a starting point.
3. Fit the model with `metab()`.

## A Maximum Likelihood Model

### 1. Choose a model structure

A model is fully specified by a concise if cryptic model name. Examples are 
"b_np_oipi_tr_plrckm.stan" and "m_np_oi_tr_plrckm.nlm". You never need to
remember these concise names; instead, you can construct and interpret such
names using the `mm_name` and `mm_parse_name` functions, respectively.

`streamMetabolizer` offers three core modeling `type`s: maximum likelihood
estimation ('mle'), hierarchical or non-hierarchical Bayesian models ('bayes'),
or nighttime regression ('night'). It is always helpful in `mm_name` to specify
the model `type` you want. Further structural details can be specified in
`mm_name` or left to the defaults.

Here are some example calls to create and parse model names:
```{r names}
three_names <- c(
  mm_name(type='mle'), # the default MLE model
  mm_name(type='mle', ode_method='euler'), # override the default ode_method
  mm_name(type='bayes')) # the default Bayesian model
three_names

# parse the above model names
mm_parse_name(three_names)
```

See the help file at `?mm_name` for a description of the arguments to `mm_name`
and results from `mm_parse_name`.

Here we will fit the default MLE model. Many others are available (see Other
Models below), but this one is common and fast. Call `mm_name` again to set the
default MLE model name/structure.
```{r mle_name}
mle_name <- mm_name(type='mle')
mle_name
```

### 2. Set the specifications

Having chosen a model, we next need to define a list of specifications for that
model. The `specs` function creates a list appropriate to the model we chose.
```{r mle_specs}
mle_specs <- specs(mle_name)
mle_specs
```

See `?specs` for definitions of all specifications. Note that most of the 
specifications in that help file are omitted from the output of
`specs(mle_name)` above - this is because MLE models are simple and don't have
many parameters to set. Any of those parameters that are included in
`model_specs` can be modified, either by calling `specs()` again or by replacing
that value in the `model_specs` list. Here is a command that sets the the inital
values of GPP, ER, and K600 for the likelihood maximization. (I've done this
just for illustration; the model results aren't affected by these particular
changes for this particular dataset.)
```{r specs_details}
mle_specs <- specs(mle_name, init.GPP.daily=2, init.ER.daily=-1, init.K600.daily=3)
```

### 3. Fit the model

Now actually fit the model using the `metab` function.
```{r mle_fit, warning=FALSE}
mle_fit <- metab(mle_specs, data=dat, info=c(site='French Creek, WY', source='Bob Hall'))
```

It's optional, but sometimes helpful, to include some sort of metadata in the
`info`, as I've done above. I've chosen to put the metadata in a character
vector, but metadata can take any format you like.

## Inspect the output

The `info` and `data` are saved in the fitted model object so we can reference
them later.
```{r info}
get_info(mle_fit)
head(get_data(mle_fit))
```

We can also get information about the model fitting process.
```{r info2}
get_fitting_time(mle_fit) # the time it took to fit the model
get_version(mle_fit) # the streamMetabolizer version used to fit the model
get_specs(mle_fit) # the specifications we passed in
```

There is a function to plot the daily metabolism estimates.
```{r plot_metab1, fig.width=7, fig.height=6}
plot_metab_preds(predict_metab(mle_fit))
```

There is also a function to plot the dissolved oxygen predictions (lines) along
with the original observations (points).
```{r plot_metab2, fig.width=7, fig.height=6}
plot_DO_preds(predict_DO(mle_fit))
```

## Other Models

Many model structures are available. We have only demonstrated one above (the
default MLE model). Here is the full list of model options:

```{r, results='asis'}
all_models <- mm_valid_names(type=c('bayes','mle','night'))
opts <- 
  bind_cols(
    tibble::tibble(model_name=all_models), 
    mm_parse_name(all_models)) %>%
  arrange(model_name)
knitr::kable(opts)
```

Remember that any of these structures can be specified by using the model name
directly or passing the corresponding arguments to `mm_name` to create the name.

## A Bayesian Model

### 1. Choose a model structure

For our second modeling example, we will specify everything about the model
structure in our call to `mm_name`. (Some of these arguments are equal to the
defaults and don't technically need to be specified, but I want to show you that
the available arguments are the same as those returned in the columns of the
above table.)
```{r bayes_name}
bayes_name <- mm_name(
  type='bayes', pool_K600='none', 
  err_obs_iid=TRUE, err_proc_acor=FALSE, err_proc_iid=TRUE, 
  ode_method='trapezoid', deficit_src='DO_mod', engine='stan')
bayes_name
```

### 2. Set the specifications

Just as for the previous model, we now pass the model name to `specs()` to get a
list of default specifications for this model. Note that the length and contents
of this new specifications list differs substantially from the list we had for
the MLE model.
```{r bayes_specs}
bayes_specs <- specs(bayes_name)
bayes_specs
```

At this point we can alter some of the specifications if desired.
```{r bayes_specs2}
# one way to alter specifications: call specs() again
bayes_specs <- specs(bayes_name, burnin_steps=100, saved_steps=200, GPP_daily_mu=3, GPP_daily_sigma=2)
# another way: use replace()
bayes_specs <- replace(bayes_specs, 'split_dates', FALSE)
```

### 3. Fit the model

We now fit the model with the new specifications. Bayesian models take longer to
run, so be patient.
```{r bayes_fit, warning=FALSE}
bayes_fit <- metab(bayes_specs, data=dat)
```

### 4. Explore as before

All of the same inspection functions are available as before, including
`get_fitting_time`, `get_info`, `predict_metab`, and so on.
```{r bayes_pred}
get_fitting_time(bayes_fit)
preds <- predict_metab(bayes_fit)
```

One slight difference is that hierarchical Bayesian models store their warnings 
and errors in attributes to the result of `predict_metab`. If there were any,
you could inspect them like this:
```{r bayes_warning}
attr(preds, 'warnings')
attr(preds, 'errors')
```

Here are the predictions from this last model:
```{r bayes_pred_tbl, results='asis'}
preds %>% 
  lapply(function(col) if(is.numeric(col)) round(col, 2) else col ) %>%
  as.data.frame() %>%
  knitr::kable()
```

You're on your way!