---
title: "Getting Started with streamMetabolizer"
author: "Alison Appling"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{getstarted}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

## Introduction

This file demonstrates how to estimate stream metabolism using example data supplied by Bob Hall for French Creek in Laramie, WY.

## Setup

### Load

Load the R libraries that will be used.
```{r libs, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(streamMetabolizer)
})
```

Load a moderately-sized example dataset from the package. Explicitly attached units are optional
but useful for confirming that the data are in the expected format. We'll use the `streamMetabolizer`
standard in defining our day to run from 4 am (`day_start=4`) to 4 am (`day_end=28`). 
```{r data}
dat <- data_metab(num_days='3', res='15', day_start=4, day_end=28, attach.units=TRUE)
```

### Explore the input data

First graphically inspect the input data.
```{r viz_inputs_DO, fig.width=7, fig.height=3}
dat %>% unitted::v() %>%
  mutate(DO.pctsat = 100 * (DO.obs / DO.sat)) %>%
  select(solar.time, starts_with('DO')) %>%
  gather(type, DO.value, starts_with('DO')) %>%
  mutate(units=ifelse(type == 'DO.pctsat', 'DO\n(% sat)', 'DO\n(mg/L)')) %>%
  ggplot(aes(x=solar.time, y=DO.value, color=type)) + geom_line() + 
  facet_grid(units ~ ., scale='free_y') + theme_bw() +
  scale_color_discrete('variable')
```

```{r viz_inputs_other, fig.width=7, fig.height=4}
labels <- c(depth='depth\n(m)', temp.water='water temp\n(deg C)', light='PAR\n(umol m^-2 s^-1)')
dat %>% unitted::v() %>%
  select(solar.time, depth, temp.water, light) %>%
  gather(type, value, depth, temp.water, light) %>%
  mutate(
    type=ordered(type, levels=c('depth','temp.water','light')),
    units=ordered(labels[type], unname(labels))) %>%
  ggplot(aes(x=solar.time, y=value, color=type)) + geom_line() + 
  facet_grid(units ~ ., scale='free_y') + theme_bw() +
  scale_color_discrete('variable')
```

### Check the input data format

Now inspect the dimensions, column names, and units of the data. Note that `unitted` data, 
such as that returned by `data_metab(..., attach.units=TRUE)`, prints out with a second row
called 'U' giving the units for each column.
```{r data_check}
dim(dat)
dat[c(1,48,96,240,288),] # some example rows
```

Our data need to have specific column names and units to fit an MLE, Bayesian, or nighttime-regression model.
To see what is required, use the `metab_inputs` function to get a 1-row example for the desired model type.
The output of metab_inputs is a 1-row unitted data.frame showing the required column names and units, 
plus one row of example data. The required inputs are identical for the model types 'mle', 'bayes', and 'night',
so here we'll just print the requriements for 'mle'.
```{r data_needs}
metab_inputs('mle', 'data')
```

Our data format matches that of the requirements from `metab_inputs`, so we're ready to go.

## Modeling Overview

There are three steps to fitting a metabolism model in `streamMetabolizer`.

1. Identify the name of the model structure you want using `mm_name()`.
2. Set the specifications for the model using defaults from`specs()` as a starting point.
3. Fit the model with `metab()`.

## A Maximum Likelihood Model

### 1. Choose a model structure

A model is fully specified by a concise if cryptic model name. 
Examples are "b_np_oipi_pm_km.stan" and "m_np_oi_pm_km.nlm". You never need
to remember these concise names; instead, you can construct and interpret 
such names using the `mm_name` and `mm_parse_name` functions, respectively. 

`streamMetabolizer` offers three core modeling `type`s: maximum likelihood estimation ('mle'), 
hierarchical or non-hierarchical Bayesian models ('bayes'), or nighttime regression ('night'). 
It is always helpful in `mm_name` to specify the model `type` you want.
Further structural details can be specified in `mm_name` or left to the defaults.

Here are some example calls to create and parse model names:
```{r names}
three_names <- c(
  mm_name(type='mle'), # the default MLE model
  mm_name(type='mle', ode_method='Euler'), # override the default ode_method
  mm_name(type='bayes')) # the default Bayesian model
three_names

# parse the above model names
mm_parse_name(three_names)
```

See the help file at `?mm_name` for a description of the arguments to `mm_name` and
results from `mm_parse_name`.

Here we will fit the default MLE model. Many others are available (see Other Models below), 
but this one is common and fast. Call `mm_name` again to set the default MLE model name/structure.
```{r mle_name}
mle_name <- mm_name(type='mle')
mle_name
```

### 2. Set the specifications

Having chosen a model, we next need to define a list of specifications for that model. 
The `specs` function creates a list appropriate to the model we chose.
```{r mle_specs}
mle_specs <- specs(mle_name)
mle_specs
```

See `?specs` for definitions of all specifications. Note that most of the 
specifications in that help file are omitted from the output of `specs(mle_name)` 
above - this is because MLE models are simple and don't have many parameters to set. 
Any of those parameters that are included in `model_specs` can be modified, either
by calling `specs()` again or by replacing that value in the `model_specs` list. 
Here is a command that sets the the inital values of GPP, ER, and K600 
for the likelihood maximization. (I've done this just for illustration; the model
results aren't affected by these particular changes for this particular dataset.)
```{r specs_details}
mle_specs <- specs(mle_name, GPP_init=2, ER_init=-1, K600_init=3)
```

### 3. Fit the model

Now actually fit the model using the `metab` function.
```{r mle_fit, warning=FALSE}
mle_fit <- metab(mle_specs, data=dat, info=c(site='French Creek, WY', source='Bob Hall'))
```

It's optional, but sometimes helpful, to include some sort of metadata in 
the `info`, as I've done above. I've chosen to put the metadata in a 
character vector, but metadata can take any format you like. 

## Inspect the output

The `info` and `data` are saved in the fitted
model object so we can reference them later.
```{r info}
get_info(mle_fit)
head(get_data(mle_fit))
```

We can also get information about the model fitting process.
```{r info2}
get_fitting_time(mle_fit) # the time it took to fit the model
get_version(mle_fit) # the streamMetabolizer version used to fit the model
get_specs(mle_fit) # the specifications we passed in
```

There is a function to plot the daily metabolism estimates.
```{r plot_metab1, fig.width=7, fig.height=6}
plot_metab_preds(predict_metab(mle_fit))
```

There is also a function to plot the dissolved oxygen 
predictions (lines) along with the original observations (points).
```{r plot_metab2, fig.width=7, fig.height=6}
plot_DO_preds(predict_DO(mle_fit))
```

## Other Models

Many model structures are available. We have only demonstrated one above (the default MLE model). 
Here is the full list of model options:

```{r, results='asis'}
all_models <- mm_valid_names(type=c('bayes','mle','night'))
opts <- 
  bind_cols(
    dplyr::data_frame(model_name=all_models), 
    mm_parse_name(all_models)) %>%
  arrange(model_name)
knitr::kable(opts)
```

Remember that any of these structures can be specified by using the model name directly or passing
the corresponding arguments to `mm_name` to create the name. 

## A Bayesian Model

### 1. Choose a model structure

For our second modeling example, we will specify everything about the model structure in our call to `mm_name`.
(Some of these arguments are equal to the defaults and don't technically need to be specified, but I want to
show you that the available arguments are the same as those returned in the columns of the above table.)
```{r bayes_name}
bayes_name <- mm_name(
  type='bayes', pool_K600='none', err_obs_iid=TRUE, err_proc_acor=FALSE, err_proc_iid=TRUE,
  ode_method='pairmeans', deficit_src='DO_mod', engine='jags')
bayes_name
```

### 2. Set the specifications

Just as for the previous model, we now pass the model name to `specs()` to get
a list of default specifications for this model. Note that the length and contents of this new
specifications list differs substantially from the list we had for the MLE model.
```{r bayes_specs}
bayes_specs <- specs(bayes_name)
bayes_specs
```

At this point we can alter some of the specifications if desired.
```{r bayes_specs2}
# one way to alter specifications: call specs() again
bayes_specs <- specs(bayes_name, burnin_steps=100, saved_steps=200, GPP_daily_mu=3, GPP_daily_sigma=2)
# another way: use replace()
bayes_specs <- replace(bayes_specs, 'split_dates', FALSE)
```

### 3. Fit the model

We now fit the model with the new specifications. Bayesian models take longer to run, so be patient.
```{r bayes_fit, warning=FALSE}
bayes_fit <- metab(bayes_specs, data=dat)
```

### 4. Explore as before

All of the same inspection functions are available as before, including `get_fitting_time`, `get_info`, `predict_metab`, and so on.
```{r bayes_pred}
get_fitting_time(bayes_fit)
preds <- predict_metab(bayes_fit)
```

One slight difference is that hierarchical Bayesian models store their warnings and errors in attributes to the result of `predict_metab`.
```{r bayes_warning}
attr(preds, 'warnings')
attr(preds, 'errors')
```

Here are the predictions from this last model:
```{r bayes_pred_tbl, results='asis'}
preds %>% 
  lapply(function(col) if(is.numeric(col)) round(col, 2) else col ) %>%
  as.data.frame() %>%
  knitr::kable()
```

You're on your way!