---
title: "Installation"
author: "Alison Appling"
date: "`r Sys.Date()`"
output:
  # rmarkdown::github_document
  rmarkdown::html_vignette:
      keep_md: true
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

# Installing streamMetabolizer

Use the `remotes` package to install `streamMetabolizer` from source.

You may first need to install the `unitted` dependency:
```{r, eval=FALSE}
remotes::install_github('USGS-R/unitted')
```

You can then install the most cutting edge version of streamMetabolizer with this command:
```{r, eval=FALSE}
remotes::install_github("USGS-R/streamMetabolizer")
```

## Software dependencies

If you plan to use Bayesian models on a Windows machine, you will need an up-to-date installation of [Rtools](https://cran.r-project.org/bin/windows/Rtools/). Run `devtools::find_rtools()` to make sure Rtools is ready to go. (Rtools is broadly useful for many R packages.) Having Rtools installed will allow you to install `rstan`, the package that `streamMetabolizer` relies on to run MCMC models.

Bayesian models require the [`rstan`](http://mc-stan.org/interfaces/rstan.html) 
interface to [Stan](http://mc-stan.org/). Sometimes this is as simple as
installing Rtools and calling the above `remotes::install_github` command, but other
times everything seems fine until you try to run a Bayesian model in
`streamMetabolizer`. Symptoms of an imperfect `rstan` installation are probably
diverse. Here's one we've seen:
```{r, eval=FALSE}
> bayes_fit <- metab(specs('bayes'), data=mydat)
Warning message:
In metab_fun(specs = specs, data = data, data_daily = data_daily,  :
  Modeling failed: argument is of length zero

> get_fit(bayes_fit)
...
$warnings
[1] "running command ''/Library/Frameworks/R.framework/Resources/bin/R' CMD config CXX 2>/dev/null' had status 1"

$errors
[1] "argument is of length zero"
```
In such cases you should refer to the detailed instructions on the `rstan` website for 
[Mac and Linux](https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Mac-or-Linux) 
or [Windows](https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Windows).

