---
title: "Installation"
author: "Alison Appling"
date: "`r Sys.Date()`"
output:
  # rmarkdown::github_document
  rmarkdown::html_vignette:
      keep_md: true
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

# Installing streamMetabolizer

To install the `streamMetabolizer` package, use the `remotes` package (running `install.packages('remotes')` first if needed). To use `remotes::install_github()` it is convenient to set a [GitHub Personal Access Token (PAT)](https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens). There are [several methods](https://usethis.r-lib.org/articles/git-credentials.html) for setting your PATs within R; the simplest is to call `Sys.setenv(GITHUB_PAT="yyyy"),
replacing yyyy with the PAT you established on the GitHub website.

You may first need to install the `unitted` dependency:
```{r, eval=FALSE}
remotes::install_github('appling/unitted')
```

You can then install the most cutting edge version of streamMetabolizer with this command:
```{r, eval=FALSE}
remotes::install_github(
  "USGS-R/streamMetabolizer", # soon to be "DOI-USGS/streamMetabolizer"
  build_vignettes = TRUE)
```

## Software dependencies for Bayesian models

The major dependency for Bayesian models is the `rstan` package, and installation of that package is rarely as simple as a call to `install.packages()`. Start at the [rstan wiki page](https://github.com/stan-dev/rstan/wiki) for the most up-to-date installation instructions, which differ by operating system.
