---
title: "Installation"
author: "Alison Appling"
date: "`r Sys.Date()`"
output:
  # rmarkdown::github_document
  rmarkdown::html_vignette:
      keep_md: true
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

# Installing streamMetabolizer

## Recommended

The most stable+current version of this package can be installed with this R command:
```{r, eval=FALSE}
install.packages("streamMetabolizer", dependencies=TRUE, 
  repos=c("https://owi.usgs.gov/R","https://cran.rstudio.com"))
```
and updated with this command:
```{r, eval=FALSE}
update.packages(oldPkgs=c("streamMetabolizer","unitted"), dependencies=TRUE,
  repos=c("https://owi.usgs.gov/R", "https://cran.rstudio.com"))
```

## For the adventurous

The in-development version of the package can be installed with the `devtools` package. 
We can make no guarantees about the stability of this version, 
but it might have new features that you'll like.
If you go this route, you will need to install the package dependencies separately, like this:
```{r, eval=FALSE}
install.packages(
  c("LakeMetabolizer","unitted","dplyr","lazyeval","lubridate","magrittr",
    "tidyr","chron","dygraphs","ggplot2","RCurl","rstan","XML","xts"),
  repos=c("https://owi.usgs.gov/R","https://cran.rstudio.com"))
```
You can then install the most cutting edge version of streamMetabolizer with this command:
```{r, eval=FALSE}
devtools::install_github("USGS-R/streamMetabolizer")
```

## Software dependencies

If you plan to use Bayesian models, you will need an up-to-date installation of [Rtools](https://cran.r-project.org/bin/windows/Rtools/). Run `devtools::find_rtools()` to make sure Rtools is ready to go. (Rtools is broadly useful for many R packages.) Having Rtools installed will allow you to install rstan, the package that `streamMetabolizer` relies on to run MCMC models.

Bayesian models require the [`rstan`](http://mc-stan.org/interfaces/rstan.html) 
interface to [Stan](http://mc-stan.org/). Sometimes this is as simple as
installing Rtools and calling the above `install.packages` or `install_github` command, but other
times everything seems fine until you try to run a Bayesian model in
`streamMetabolizer`. Symptoms of an imperfect `rstan` installation are probably
diverse. Here's one we've seen:
```{r, eval=FALSE}
> bayes_fit <- metab(specs('bayes'), data=mydat)
Warning message:
In metab_fun(specs = specs, data = data, data_daily = data_daily,  :
  Modeling failed: argument is of length zero

> get_fit(bayes_fit)
...
$warnings
[1] "running command ''/Library/Frameworks/R.framework/Resources/bin/R' CMD config CXX 2>/dev/null' had status 1"

$errors
[1] "argument is of length zero"
```
In such cases you should refer to the detailed instructions on the `rstan` website for 
[Mac and Linux](https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Mac-or-Linux) 
or [Windows](https://github.com/stan-dev/rstan/wiki/Installing-RStan-on-Windows).

