---
title: "Installation"
author: "Alison Appling"
date: "`r Sys.Date()`"
output:
  # rmarkdown::github_document
  rmarkdown::html_vignette:
      keep_md: true
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

# Installing streamMetabolizer

Use the `remotes` package to install `streamMetabolizer` from source.

You may first need to install the `unitted` dependency:
```{r, eval=FALSE}
remotes::install_github('appling/unitted')
```

You can then install the most cutting edge version of streamMetabolizer with this command:
```{r, eval=FALSE}
remotes::install_github("USGS-R/streamMetabolizer")
```

## Software dependencies for Bayesian models

The major dependency for Bayesian models is the `rstan` package, and installation of that package is rarely as simple as a call to `install.packages()`. Start at the [rstan wiki page](https://github.com/stan-dev/rstan/wiki) for the most up-to-date installation instructions, which differ by operating system.
