---
title: "Model Structures"
author: "Alison Appling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=80)
library(streamMetabolizer)
```

# Basic model types in streamMetabolizer

Many model structures are available in the streamMetabolizer package. These fall into three general `type`s:

* `bayes` - Inverse Bayesian modeling of GPP, ER, and K600
* `mle` - Inverse modeling by maximum likelihood estimation of GPP, ER, and optionally K600
* `night` - Nighttime regression for estimation of K600 and ER

Two additional model `type`s can be useful in specific situations:

* `Kmodel` - Builds a multi-day relationship between K600 and predictor variables; can be used in conjunction with `mle` and/or `night`
* `sim` - Simulates dissolved oxygen "data" for model testing

# Specific model structure

A complete specification of a model structure begins with a basic `type` but
also includes other information about the model error assumptions, any
hierarchical components, the numerical integration method, the process equations
used to describe GPP and ER, and more. A model structure is fully specified by a
concise if cryptic model name. Examples are "b_np_oipi_tr_plrckm.stan" and
"m_np_oi_tr_plrckm.nlm". You never need to remember these concise names;
instead, you can construct and interpret such names using the `mm_name` and
`mm_parse_name` functions, respectively.

# Setting a model name

The `mm_name()` function creates a cryptic (but complete and concise) model name from several options that you can specify. For example:

```{r}
mm_name(type='bayes') # the default Bayesian model
mm_name(type='bayes', pool_K600='normal') # a Bayesian model with simple pooling of K600 to a shared mean
mm_name(type='mle') # the default MLE model
mm_name(type='mle', ode_method='euler') # an MLE model with a simpler ode_method

```

See the help file at `?mm_name` to learn about the structural options that you can adjust.

# Interpreting a model name

The `mm_parse_name()` function splits a cryptic model name into its component features. For example:

```{r}
mm_parse_name(c('b_np_oipi_tr_plrckm.stan', 'm_np_pi_tr_psrqkm.nlm'))
```

See the help file at `?mm_parse_name`, and refer again to `?mm_name` for definitions of the column names in the output of `mm_parse_name`.

# Listing all the options

You can see the full list of available model structures by 
calling `mm_valid_names()` (here we'll just print a sampling of the full list):

```{r}
valid_names <- mm_valid_names(type=c('bayes','mle','night'))
length(valid_names)
c(valid_names[seq(1,length(valid_names),length.out=20)], '...')
```

# Using a model name

Once you have a model name in hand, pass that name to `metab()` to fit a model with the selected structure. See the [Quickstart](get_started.html) tutorial for more.
