---
title: "Bayesian Models"
author: "Alison Appling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This page provides details on how to get the best results from a Bayesian metabolism model. See [Quickstart](get_started.html) for an example of preparing and fitting a Bayesian model.

## Why Bayesian models?

Bayesian models are slower than the alternatives, but they offer several advantages:
* They can be state-space models, including both observation error and process error.
* They can be hierarchically structured, allowing you to use information from many days of data to inform estimates on each individual day.
* They produce more accurate and nuanced uncertainty estimates.

## Configuring Bayesian models

The structure and specifications of Bayesian models require special attention. For more information, see the help file at `?specs`, especially the `Relevant arguments` section and the parameter definitions. Also see the vignette called 'fit_fix_k'.

## Inspecting Bayesian models

### Stan MCMC information

After fitting a `streamMetabolizer` model, you can pull out the original Stan model object using `get_mcmc(mm)` where `mm` is your fitted metabolism model. You'll need the `rstan` package to inspect the outputs from there, so run `library(rstan)` before proceeding.

See `?stanfit` for some ideas on how to inspect the Stan model object. My go-to first method is `traceplot()` (and the `inc_warmup=TRUE` arg to `traceplot()` can be handy). `plot()`, `print()`, and `summary()` are also useful. 

### Model warnings and errors

Unlike other model types in streamMetabolizer, Bayesian models sometimes have
overall warnings and errors not specific to any one day. If there are any, you
will see a note in the 'warnings' or 'errors' columns of the model printout, and you can
see the full message[s] as elements in the list returned by `get_fit()`. There's
also a [smaller] possibility of errors or warnings directly associated with
metabolism prediction rather than fitting, inspectable in the 'warnings' and
'errors' columns of the data.frame returned by `predict_metab()`.

```{r bayes_warning, eval=FALSE}
# here's where you'd find fitting messages:
select(get_params(mm), warnings, errors)
get_fit(mm)$warnings
get_fit(mm)$errors

# and prediction messages
select(predict_metab(mm), warnings, errors)
```
