---
title: "MLE Models"
author: "Alison Appling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MLE Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file provides an example of fitting an MLE (maximum likelihood estimation) model, following the general outline in the [Quickstart](get_started.html) tutorial.

As a reminder, there are four steps to fitting a metabolism model in `streamMetabolizer`.

1. Prepare and inspect the input data.
2. Choose a model configuration appropriate to your data.
3. Fit the model.
4. Inspect the output.

# Preliminaries

If you haven't already installed the package, see the [Installation](installation.html) tutorial.

Next load the R libraries. Only `streamMetabolizer` is required to run models, but we'll also be using `dplyr` to inspect the results.
```{r libs, warning=FALSE, message=FALSE}
library(streamMetabolizer)
library(dplyr)
```

# 1. Preparing the input data

Load a small example dataset from the package (data are from French
Creek in Laramie, WY, courtesy of Bob Hall). Explicitly attached units are
optional but useful for confirming that the data are in the expected format.
We'll use the `streamMetabolizer` standard in defining our day to run from 4 am
(`day_start=4`) to 4 am (`day_end=28`).
```{r data}
dat <- data_metab(num_days='3', res='15', day_start=4, day_end=28, attach.units=TRUE)
```

# 2. Configuring the model

There are two steps to configuring a metabolism model in `streamMetabolizer`.
 
a. Identify the name of the model structure you want using `mm_name()`.
b. Set the specifications for the model using defaults from`specs()` as a starting point.

## 2a. Choose a model structure

Call `mm_name` to choose a specific MLE model name/structure. Here we will fit
the default MLE model. Many others are available (see [Model Structures](model_structures.html)
and [GPP and ER equations](gpp_er_eqs.html)), but this one is common and fast.
```{r mle_name}
mle_name <- mm_name(type='mle')
mle_name
```

## 2b. Set the specifications

Having chosen a model, we next need to define a list of specifications for that
model. The `specs` function creates a list appropriate to the model we chose.
```{r mle_specs}
mle_specs <- specs(mle_name)
mle_specs
```

See `?specs` for definitions of all specifications. Note that most of the 
specifications in that help file are omitted from the output of
`specs(mle_name)` above - this is because MLE models are simple and don't have
many parameters to set. Any of those parameters that are included in
`mle_specs` can be modified, either by calling `specs()` again or by replacing
that value in the `mle_specs` list. Here is a command that sets the the inital
values of GPP, ER, and K600 for the likelihood maximization. (I've done this
just for illustration; the model results aren't affected by these particular
changes for this particular dataset, and you will seldom need to edit these values.)
```{r specs_details}
mle_specs <- specs(mle_name, init.GPP.daily=2, init.ER.daily=-1, init.K600.daily=3)
```

# 3. Fitting the model

Now actually fit the model using the `metab` function.
```{r mle_fit, warning=FALSE}
mm <- metab(mle_specs, data=dat, info=c(site='French Creek, WY', source='Bob Hall'))
```

It's optional, but sometimes helpful, to include some sort of metadata in the
`info`, as I've done above. I've chosen to put the metadata in a character
vector, but metadata can take any format you like.

## 4. Inspecting the model

Models show lots of relevant information if you simply print them at the command line.
```{r show}
mm
```

You can also extract specific pieces of information using designated accessor
functions. For example, the `info` and `data` are saved in the fitted model
object and can be pulled out with `get_info` and `get_data`, respectively.
```{r info}
get_info(mm)
head(get_data(mm))
```

We can also get information about the model fitting process.
```{r info2}
get_fitting_time(mm) # the time it took to fit the model
get_version(mm) # the streamMetabolizer version used to fit the model
get_specs(mm) # the specifications we passed in
```

There is a function to plot the daily metabolism estimates.
```{r plot_metab1, fig.width=7, fig.height=4.5}
plot_metab_preds(mm)
```

There is also a function to plot the dissolved oxygen predictions (lines) along
with the original observations (points).
```{r plot_metab2, fig.width=7, fig.height=6}
plot_DO_preds(mm)
```

You can output the daily and instantaneous predictions to data.frames for
further inspection.
```{r pred_dfs}
predict_metab(mm)
head(predict_DO(mm))
```
