---
title: "Two-Station Models"
author: "Two-Station Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-Station Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=80)
```

## Background

The two-station approach estimates reach-scale gross primary production
(GPP), ecosystem respiration (ER), and the gas exchange rate constant (K600,
d^-1^) by pairing an upstream and a downstream dissolved oxygen sensor separated by a known
travel time. Rather than modeling O~2~ dynamics at a single point, the
method tracks how a parcel of water changes in O~2~ content and O~2~ percent 
saturation as it travels between the two stations — making it appropriate for reaches 
where O~2~ is not at steady state (e.g., dam
tailwaters, reaches below waterfalls, or reaches with significant tributary inputs 
or hyporheic exchange).

The implementation in `streamMetabolizer` follows the variable-flow
two-station (VFTS) method of Bishop et al. (2026), which accounts for
discharge-driven variation in travel time. The model can be used for rivers and
streams with no sub-daily changes in flow as well as for systems that experience
significant sub-daily flow variability (such as results from hydropeaking). 
The model is fit with Stan and follows the same `specs()` → `metab()` → `predict_metab()` 
workflow as the one-station Bayesian models (see the [Quickstart](get_started.html) and
[Bayesian Models](models_bayes.html) tutorials).

**When to use the two-station model:**

- The study reach has known spatial heterogeneity such that dissolved oxygen
  is not at steady state within the reach (large tributary inputs, hyporheic zones, 
  impoundments, waterfalls, wastewater inputs)
- The study reach experiences event-based inputs such that dissolved oxygen
  is not at steady state within the reach (e.g., monsoon or wildfire runoff)
 
**Additional requirements of the two-station model:**

- An estimate is available for the travel time between sensors
- Paired sensors should be placed within the 80% turnover 
  reach for oxygen (ideal distance is ~0.7v/K600, and should not fall below 0.4v/K600; 
  Reichert et al. 2009; Demars et al. 2015). 

<!-- TODO: a decision checklist or flowchart for choosing between one- and
two-station models could go here. -->

**When to use a one-station model instead:**

- The reach is well mixed and spatially homogeneous
- A state space model structure that incorporates both process error and 
  observation error is needed (the two-station model offers
  process error only; see [Known limitations](#known-limitations))
- An autocorrelated error structure is required

```{r libs, warning=FALSE, message=FALSE}
library(streamMetabolizer)
library(dplyr)
```

## Getting your data in

The `metab_bayes_2s()` requires a single formatted data file.  There are two
ways the user can approach compiling this file. In path A, the user can start with the 
raw shape (recommended for most users). Path A is the better starting point when 
your data come off the sensors as separate series, which is the usual case. 
In path B, the user has already assembled the data, and starts with the formatted shape 
(recommended for advanced users only, particularly when 
an alternate method for estimating fractional light is desired).

**A. Raw.** You supply three separate data.frames — upstream sensor,
downstream sensor, and light — to `mm_format_data_2s()`, which joins them
onto a common time grid, bridges short gaps, converts light to the fractional 
form the model needs, and returns a formatted data.frame (described below B. Formatted).
`two_station_raw_example` shows that shape.

**B. Formatted.** You supply one data.frame with all nine required
columns, already joined onto a single time grid. `two_station_example`
shows that shape.  
 

### The raw shape

`two_station_raw_example` is a list of three data.frames — the same Glen
Canyon reach as `two_station_example`, but as it comes off the sensors: not
joined, not aligned to a common grid, and with light still as raw PAR.  The
units for light are unimportant as long as they are consistent, since they 
will be used to calculate fractions of total daily light.

```{r ts_raw_data}
data(two_station_raw_example)
lapply(two_station_raw_example, function(df) head(df, 2))
```

Each input needs package-convention column names: `timestamp` (POSIXct, UTC)
plus `DO.obs` and `DO.sat` for `upstream`; `timestamp`, `DO.obs`, `DO.sat`,
`temp.water`, `depth`, and `travel.time` for `downstream`; and `timestamp`
plus `light` for `light`. Renaming columns to match is a step you do yourself, 
before calling `mm_format_data_2s()`.  The light data.frame should be reach averaged
light corresponding to the downstream time stamp.

`DO.sat` for each station can be computed from that station's
temperature and barometric pressure with `calc_DO_sat()`. We urge the user
to be cautious about using raw sensor estimates of oxygen saturation as 
these can be highly sensitive to the barometric pressure calibration on 
your instrument.

A quick demonstration with simulated values:

```{r ts_do_sat_demo}
set.seed(47)
temp.water <- round(runif(20, min = 10, max = 25), 1)
pressure.air <- round(runif(20, min = 980, max = 1010), 1)

DO_sat_demo <- calc_DO_sat(temp.water, pressure.air)
head(DO_sat_demo)
```

`pressure.air` is barometric pressure in millibars. 

```{r ts_format}
formatted <- mm_format_data_2s(
  upstream   = two_station_raw_example$upstream,
  downstream = two_station_raw_example$downstream,
  light      = two_station_raw_example$light)

head(formatted, 3)
```

`mm_format_data_2s()` does four things: it trims upstream and downstream to
their shared deployment window (reporting what it dropped), snaps all three
series onto a common timestep grid so that sensors logging a few minutes
apart still match, interpolates short gaps, and converts raw light into the
within-day proportion. The message above reports the gap filling.

`downstream` is the backbone: every downstream row survives into the output,
carrying `NA` wherever no upstream or light observation matched it.

### The formatted shape (produced by mm_format_data_2s())

```{r ts_example_data}
data(two_station_example)
head(two_station_example, 3)
```

| Column | Description |
|---|---|
| `solar.time` | Mean solar time (POSIXct, UTC tzone), on a regular timestep, labelling the downstream observations |
| `DO.obs.up` | Upstream observed dissolved oxygen (mgO2 L^-1^) |
| `DO.sat.up` | Upstream O~2~ saturation (mgO2 L^-1^) |
| `DO.obs.down` | Downstream observed dissolved oxygen (mgO2 L^-1^) |
| `DO.sat.down` | Downstream O~2~ saturation (mgO2 L^-1^) |
| `light` | Within-day light proportion (unitless, 0–1) |
| `depth` | Reach-averaged depth (m) — not point depth |
| `temp.water` | Water temperature at the downstream station (degC) |
| `travel.time` | Reach travel time between stations (d) |

This formatted shape matches each downstream observation to the upstream
water parcel that became it, using the travel times provided.

`mm_format_data_2s()` does not convert timestamps to mean solar time. As in
the one-station models, `solar.time` is expected to already be mean solar
time (POSIXct, UTC tzone); convert your raw timestamps first with
`calc_solar_time()` or `convert_UTC_to_solartime()` (see the
[Data Preparation](data_prep.html) tutorial). Approximate local,
non-daylight-saving clock time is a workable substitute, but mean solar time
aligns the 06:00-06:00 day window and the within-day light proportion to the
true diel light cycle.

Three columns deserve attention relative to one-station data.

**`depth`** is the reach-averaged depth between the two stations, not the
point depth at either sensor. Using point depth will bias the metabolism
estimates.

**`travel.time`** is in days — a 6-hour travel time is `0.25`. It may vary
from row to row, which is the point of the variable-flow method: as
discharge changes, so does the time a parcel of water spends in the reach.
A constant travel time is equally valid and simply repeats one value down
the column.

**`light`** is *not* raw photosynthetically active radiation, unlike the
`light` column in the one-station models. It is the within-day proportion of
light: the light reaching the reach over each row's upstream travel-time
window, divided by that day's total. It is a unitless fraction between 0 and
1. Nothing checks this — supplying raw PAR in this column does not raise an
error, it just produces meaningless estimates — so getting it right is on
you when hand-formatting. If you use `mm_format_data_2s()`, the conversion
is done for you from raw light.

<!-- TODO: consider validating that hand-supplied light is in [0, 1] and
erroring on out-of-range (raw PAR) input. -->


## How a two-station day is defined

In streamMetabolizer, two-station day runs from **06:00 to 06:00 the next day**
and this time window is not changeable by the user (e.g. there is no 
`day_start`/`day_end` argument for two-station models. The two-station model
partitions the record into non-overlapping 24-hour days. 

To see which calendar dates your record covers, derive a day label from
`solar.time` — shifting back by the 06:00 day start — and tabulate it:

```{r ts_day_label}
# unitted::v() strips the units the package data carry, so that plain
# date arithmetic works
solar_time <- unitted::v(formatted$solar.time)
day_label <- as.Date(solar_time - as.difftime(6, units='hours'), tz='UTC')

head(unique(day_label))
```

A day is modeled only if it is **completely** covered — every timestep from
06:00 through 05:45 the next morning (96 rows at a 15-minute timestep).
Completeness is assessed *after* short gaps are bridged (see **Gaps in the
record** below), so a day with only a brief dropout still counts as complete.
A day left incomplete by a gap too long to bridge is dropped rather than
modeled on partial data.

### Lead-in data

Each modeled row's upstream value is taken from an earlier row: the parcel
of water arriving downstream at a given timestamp left the upstream station
`travel.time` earlier. So the first stretch of your dataset serves as
*lead-in* — those rows supply upstream values for later rows without being
modeled themselves.

The practical consequence: **start your data before the first day you want
estimates for**, by at least the longest travel time in the record. Rows
lacking lead-in are not an error; they simply do not get their own estimates,
and the (partial) day they belong to is dropped.

`two_station_example` comes with a lead-in block already prepended to the
start of its six-year record. In the [worked example](two_station_fitting.html)
we instead create one by starting the subset 12 hours before the first day
of interest.

See [Two-Station Models: Estimating Travel Time](two_station_travel_time.html)
for how travel time is estimated in practice, including the travel-time
ceiling and how uncertainty in travel time affects metabolism estimates.

## Gaps in the record

Short interruptions — a sonde that logged nothing for a few timesteps, or
logged `NA` — are bridged by linear interpolation, so that a day marred by a
brief dropout can still be modeled instead of discarded. The tolerance is
`specs$max_gap_hours`, which defaults to **1 hour** and may be set as high as
2. Longer runs are left in place, and the days containing them are dropped.

A few details worth knowing:

- Missing timesteps and `NA` values are treated the same way, and a gap is
  measured by the time it spans, not the number of rows missing across it.
- Gaps at the very start or end of the record are never filled, since that
  would mean extrapolating from one side.
- Interpolated timesteps are part of the modeled record, so `predict_DO()`
  returns predictions of them, and the predicted `DO.obs.down` at those
  timesteps is the interpolated value rather than an observation.
- If you prepared data with `mm_format_data_2s()`, gaps are already filled
  and nothing more happens at fitting time. Pass the same `max_gap_hours` to
  both if you change it from the default.

Gap filling runs on both paths, but it is cleaner on raw data. On the DO
columns, linear interpolation is straightforward either way. On `light` it
is not: once light has been converted to the within-day proportion,
interpolating it cannot repair that day's total, which is itself short by
the missing terms. `mm_format_data_2s()` avoids this by filling gaps while
`light` is still raw, before the conversion. If you hand-format, fill gaps
in your raw series before computing the light proportion.

### Why a day may be missing from your results {#missing-days}

It is normal for a fitted two-station model to report **fewer days than your
input data appears to span**. There are four distinct reasons, and it is
worth knowing which one applies before concluding something went wrong.

1. **Incomplete 06:00–06:00 coverage.** The day does not fill its window.
   This is what happens at the edges of a dataset whose bounds don't fall on
   06:00 — including, by design, the lead-in block at the start.
2. **A gap too long to bridge.** A run of missing observations longer than
   `max_gap_hours` remains in the record, leaving the day incomplete.
3. **Travel time over the ceiling.** The day's longest `travel.time` exceeds
   `max_travel_time_hours`.
4. **No observations at all.** If a whole day is absent from the input —
   sensors offline for days, say — there is nothing to report on, and the
   day simply never appears.

Cases 1 and 3 produce a message naming the days and the reason:

```
dropping 1 day(s) that do not fill the 06:00-06:00 window: 2009-08-23 (19 of 96 expected observations)
```

```
dropping 6 day(s) whose travel.time exceeds the 10-hour ceiling: 2009-08-23 (10.42 hours), 2009-08-24 (11.03 hours), ...
```

Case 2 surfaces as case 1: the gap is what makes the day incomplete, so the
message you see names the day and its observation count. Case 4 is the one
that is genuinely silent — a day with no rows is never a candidate in the
first place, so nothing is reported about it.

To reconcile input days against fitted days, label your input rows and
compare:

```{r ts_missing_days}
obs_per_day <- table(day_label)

length(obs_per_day)          # days represented in the input
sum(obs_per_day == 96)       # days fully covered at a 15-minute timestep
```

Across the full raw example, `r length(obs_per_day)`
calendar days are represented but only `r sum(obs_per_day == 96)`
are complete. Checking this before fitting is quicker than inferring it from the results.

If *no* complete days survive, `metab_bayes_2s()` raises an error rather than
returning an empty model.

See [Two-Station Fitting and Worked Example](two_station_fitting.html) for a
hands-on walk-through, including a worked example and priors/limitations.


### References

Bishop, I.W., Deemer, B.R., Kennedy, T.A., Payn, R.A., Hall Jr, R.O.
and Yackulic, C.B., 2026. A simplified two-station approach for modeling
metabolism in dam tailwaters subject to diel flow variation. *Limnology
and Oceanography: Methods*, p.e70066.
https://doi.org/10.1002/lom3.70066

Demars, B. O. L., J. Thompson, and J. R. Manson. 2015.
“Stream Metabolism and the Open Diel Oxygen Method:
Principles, Practice, and Perspectives.” Limnology and Ocean-
ography: Methods 13, no. 7: 356–374. https://doi.org/10.
1002/lom3.10030.

Reichert, P., U. Uehlinger, and V. Acuna. 2009. Estimating stream metabolism from 
oxygen concentrations: effect of spatial heterogeneity. *Journal of Geophysical 
Research- Biogeosciences*, 114, no. G3; 2008JG000917. https://doi.org/10.
1029/2008JG000917.
