---
title: "Two-Station Models: Fitting and Worked Example"
author: "Two-Station Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-Station Models: Fitting and Worked Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=80)
```

```{r libs, warning=FALSE, message=FALSE}
library(streamMetabolizer)
library(dplyr)
```

This vignette continues from [Two-Station Models](two_station.html),
covering configuration, fitting, and worked examples using the two-station
model.

## Worked example

We'll fit two short windows drawn from `two_station_example`, chosen to each
demonstrate one of the two most common reasons a day gets dropped: a
window-incomplete drop, and a `day_tests`/NA-validation drop. Both windows
are 14 days with a 1-day lead-in buffer before the window start, matching
the buffer convention used throughout this vignette.

**Window A** — 2011-06-29 through 2011-07-12, with lead-in starting
2011-06-28. This window's longest travel time is under 6.5 hours,
comfortably inside the 10-hour default ceiling.

**Window C** — 2011-10-18 through 2011-10-31, with lead-in starting
2011-10-17. Travel time here stays under 6 hours, also well inside the
ceiling.

```{r ts_windows}
window_a_start <- as.POSIXct("2011-06-28 00:00:00", tz="UTC")
window_a_end   <- as.POSIXct("2011-07-13 00:00:00", tz="UTC")  # exclusive

window_c_start <- as.POSIXct("2011-10-17 00:00:00", tz="UTC")
window_c_end   <- as.POSIXct("2011-11-01 00:00:00", tz="UTC")  # exclusive

dat_a <- two_station_example[two_station_example$solar.time >= window_a_start &
                                two_station_example$solar.time < window_a_end, ]
dat_c <- two_station_example[two_station_example$solar.time >= window_c_start &
                                two_station_example$solar.time < window_c_end, ]
```

### Configuring the model

`mm_name()` returns the model structure. Unlike the one-station types,
`'bayes_2s'` must be spelled out in full — it cannot be abbreviated, because
`'bayes'` is a prefix of it.

```{r ts_name}
bayes_2s_name <- mm_name('bayes_2s')
bayes_2s_name
```

There is a single two-station model structure, so there are no
`pool_K600`, `err_*`, etc. choices to make here as there are for
`mm_name('bayes')` (see issue #441).

<!-- TODO(#441): additional bayes_2s model structures and K600 pooling
options will make these choices meaningful. -->

`specs()` supplies the defaults; then you can use revise() to specify your own specs:

```{r ts_specs}
bayes_2s_specs <- specs(bayes_2s_name)
bayes_2s_specs <- revise(bayes_2s_specs, burnin_steps=500, saved_steps=500,
                         n_chains=4, n_cores=4)
```

### Fitting: a window-incomplete drop

```{r ts_fit_a, eval=FALSE}
mm_a <- metab_bayes_2s(bayes_2s_specs, data=dat_a)
```

```
dropping 1 day(s) that do not fill the 06:00-06:00 window: 2011-06-29 (78 of 96 expected observations)
```

The lead-in day (2011-06-28) is set aside as expected, supplying upstream
values for 2011-06-29 without being modeled itself. But 2011-06-29 *also*
drops — it follows a real gap in the sensor record (2011-06-28 has zero
rows for part of the day), and only 78 of the 96 expected observations
survive into its travel-time window. Thirteen of the fourteen requested
days come out fitted; the fourteenth is flagged, not silently absent,
more on that below.

`metab()` dispatches to `metab_bayes_2s()` on the model name, so
`metab(bayes_2s_specs, data=dat_a)` is equivalent and lets the two-station
model drop into the same chained workflow as the rest of the package.

### Fitting: a `day_tests` drop

```{r ts_fit_c, eval=FALSE}
mm_c <- metab_bayes_2s(bayes_2s_specs, data=dat_c)
```

Window C's lead-in day sets aside cleanly, and all thirteen remaining days
in the window fill their travel-time windows completely — but 2011-10-31
fails a different check. The upstream sensor stops reporting partway
through the day, leaving `NA` values in `DO.obs.up` and `DO.sat.up` that
make the day unfittable even though the window itself is otherwise
complete. This is the case a plot makes obvious: the downstream trace
keeps recording normally while the upstream trace cuts out mid-day, right
before the next real gap in the record begins on 2011-11-01. 

### Inspecting the results

`predict_metab()` gives daily GPP, ER, and K600 with credible intervals.
Both dropped days appear as their own row, with every numeric column `NA`:

```{r ts_pred_metab_a, eval=FALSE}
predict_metab(mm_a)%>%
  select(date,GPP,ER,K600,msgs.fit,warnings,errors)
```

```
# A tibble: 14 × 13
   date         GPP GPP.lower GPP.upper     ER ER.lower ER.upper  K600 K600.lower
   <date>     <dbl>     <dbl>     <dbl>  <dbl>    <dbl>    <dbl> <dbl>      <dbl>
 1 2011-06-29    NA        NA        NA     NA       NA       NA    NA         NA
 2 2011-06-30  9.53      9.14      9.92 -10.8    -11.9    -9.79  15.8       14.6
 3 2011-07-01  7.68      7.37      7.97  -7.44    -8.37   -6.49  10.4        9.15
 4 2011-07-02  8.80      8.44      9.16  -8.29    -9.20   -7.35  11.2        9.96
 5 2011-07-03  8.27      7.98      8.56  -7.23    -7.90   -6.57  10.7        9.72
# ℹ 9 more rows
# ℹ 4 more variables: K600.upper <dbl>, msgs.fit <chr>, warnings <chr>,
#   errors <chr>
```

The `msgs.fit` column on a dropped day carries a fixed-width status code
rather than the reason text — for window A's 2011-06-29 row it reads as
six blank spaces followed by `"E"`. `warnings` and `errors` in
`predict_metab()` output stay blank even for dropped days; the actual
reason lives in `get_params()` and `get_fit()`, not here.

`get_params()` reports the same quantities with standard deviations, and
this is where the reason text actually shows up:

```{r ts_get_params, eval=FALSE}
get_params(mm_a)
get_params(mm_c)
```

```
# Window A, row 1:
  date        GPP.daily GPP.daily.sd ER.daily ER.daily.sd K600.daily K600.daily.sd warnings errors
  2011-06-29         NA           NA       NA          NA         NA            NA       ""  "does not fill the 06:00-06:00 window (78 of 96 expected observations)"

# Window C, row 14:
  date        GPP.daily GPP.daily.sd ER.daily ER.daily.sd K600.daily K600.daily.sd warnings errors
  2011-10-31         NA           NA       NA          NA         NA            NA       ""  "NAs in DO.obs.up; NAs in DO.sat.up"
```

`predict_DO()` returns the modeled downstream oxygen. Note the column
names: **`DO.obs.down` and `DO.mod.down`**, not the one-station `DO.obs`
and `DO.mod`. Code written against one-station output will need adjusting.

Convergence diagnostics live in `get_fit()`. For two-station models the
`Rhat` columns are per-day, in the `daily` element, which also carries the
`valid_day` flag — the only place that column appears:

```{r ts_valid_day, eval=FALSE}
get_fit(mm_a)$daily %>%
  select(date, valid_day, errors)
```

```
# A tibble: 14 × 3
   date       valid_day errors
   <date>     <lgl>     <chr>
 1 2011-06-29 FALSE     "does not fill the 06:00-06:00 window (78 of 96 expected observations)"
 2 2011-06-30 TRUE      ""
 3 2011-07-01 TRUE      ""
# ℹ 11 more rows
```

`get_fit()` also carries `warnings` and `errors` elements for messages not
tied to any single day, and per-day `warnings`/`errors` columns within
`daily`, as described in [Bayesian Models](models_bayes.html).

`get_mcmc()` returns the underlying `stanfit` object for inspection with
`rstan`:

```{r ts_mcmc, eval=FALSE}
mcmc <- get_mcmc(mm_a)
rstan::traceplot(mcmc, pars='K600_daily', nrow=3)
```

## Fitting jointly or one day at a time

By default every date is fit together in one Stan call. Setting
`split_dates=TRUE` fits each date in its own call instead:

```{r ts_perday, eval=FALSE}
perday_specs <- revise(bayes_2s_specs, split_dates=TRUE)
mm_a_perday <- metab_bayes_2s(perday_specs, data=dat_a)
get_params(mm_a_perday)
```

```
# A tibble: 13 × 9
   date       GPP.daily GPP.daily.sd ER.daily ER.daily.sd K600.daily
   <date>         <dbl>        <dbl>    <dbl>       <dbl>      <dbl>
 1 2011-06-30      9.54        0.178   -10.85       0.460       15.86
 2 2011-07-01      7.69        0.161    -7.47       0.446       10.47
 3 2011-07-02      8.81        0.176    -8.31       0.436       11.20
 4 2011-07-03      8.28        0.148    -7.25       0.331       10.71
 5 2011-07-04      6.87        0.121    -1.99       0.238        4.39
# ℹ 8 more rows
# ℹ 3 more variables: K600.daily.sd <dbl>, warnings <chr>, errors <chr>
```

Now we only have 13 rows, not 14 — the flagged day (2011-06-29) never reaches Stan
in per-day mode, since each date is fit in its own separate call rather
than joining the pooled fit.

The difference that matters is the error term. The Stan model
carries one `sigma` shared across whatever dates it is given, so fitting
one day at a time estimates a separate `sigma` per day rather than one
pooled across the record. Joint fitting is the default because the pooled
estimate uses the whole record.

Point estimates track closely between the two modes — GPP, ER, and K600
daily means differ by only a few percent day to day. What moves is
uncertainty, and not uniformly: for 2011-07-06, the per-day `ER.daily.sd`
(0.241) is less than half the joint fit's (0.546); for 2011-07-08, per-day
is slightly *higher* than joint (0.964 vs. 0.938). Which direction it
moves depends on how much that day's own data supports a tight estimate
versus how much it was relying on the pooled `sigma` — there's no single
rule of thumb for which mode gives tighter intervals on a given day.

Fitting one day at a time does not mean slicing the data by date. Each
day's upstream observations still come from earlier rows, across the day
boundary, exactly as in the joint fit.

One consequence for `get_mcmc()`: in joint mode it returns a single
`stanfit`; in per-day mode it returns a list of `stanfit` objects named by
date — one entry per *fitted* day, so a dropped day has no entry at all.

```{r ts_perday_mcmc, eval=FALSE}
str(get_mcmc(mm_a_perday), max.level=1, list.len=3)
```

```
List of 13
 $ 2011-06-30:Formal class 'stanfit' [package "rstan"] with 10 slots
 $ 2011-07-01:Formal class 'stanfit' [package "rstan"] with 10 slots
 $ 2011-07-02:Formal class 'stanfit' [package "rstan"] with 10 slots
  [list output truncated]
```

## Fitting the pre-formatted example

`two_station_example` is already formatted and needs no preparation — it
goes straight into the model. Here we fit just the first 14 days for a
quick, build-friendly demo:

```{r ts_example_fit, eval=FALSE}
dat_start <- two_station_example[1:1344, ]
mm_example <- metab_bayes_2s(specs(mm_name('bayes_2s')), data=dat_start)
nrow(get_params(mm_example))
```

## Priors

The two-station model takes six priors through `specs()`.

**GPP and ER** use the same defaults as the one-station Bayesian models —
`GPP_daily_mu` (3.1), `GPP_daily_sigma` (6.0), `ER_daily_mu` (-7.1), and
`ER_daily_sigma` (7.1). There are no two-station-specific versions of these
to look for.

**K600** is set through `k600_velocity_lnorm_meanlog` (default `log(3.48)`,
about 1.247) and `k600_velocity_lnorm_sdlog` (default 0.5) — the meanlog and
sdlog of a lognormal prior. The names differ from the one-station
`K600_daily_meanlog`/`K600_daily_sdlog` because the quantity differs: the
two-station model puts its prior on the gas exchange velocity k600
(m d^-1^), which the Stan model divides by each timestep's depth. The
daily rate-scale `K600_daily` (d^-1^) reported in the results is a derived
quantity, computed for comparability with the one-station output.

These K600 defaults are generic, not tuned to any site, and K600 in
particular is worth revisiting for your own reach. Independent estimates,
from a tracer study, or from one-station fits on a hydraulically similar
reach, make a better-centered prior; `qlnorm()` will tell you what
quantiles a candidate `meanlog`/`sdlog` pair implies. (`plot_distribs()`
can visualize the `GPP_daily` and `ER_daily` priors from a two-station
specs list, but does not recognize the `k600_velocity_lnorm_*` names.)

```{r ts_priors, eval=FALSE}
site_specs <- revise(bayes_2s_specs,
                     k600_velocity_lnorm_meanlog=log(15), k600_velocity_lnorm_sdlog=0.7,
                     GPP_daily_mu=8, GPP_daily_sigma=4)
```

Note that K600 here is reaeration over the reach *between* the two stations,
a reach-integrated quantity. There is a single K600 estimated per day; no
hierarchical or binned pooling across days is available for two-station
models (see issue #441).

<!-- TODO(#441): hierarchical / binned K600 pooling for bayes_2s. -->

## Known limitations {#known-limitations}

- **No K600 pooling.** K600 is estimated independently for each day. The
  hierarchical and binned pooling options offered by one-station models are
  not available currently.
- **`plot_DO_preds()` does not work on two-station models.** It expects the
  one-station `DO.obs`/`DO.mod` column names. Plot `predict_DO()` output
  directly for now. `plot_metab_preds()` does work.
- **`predict_DO(use_saved=FALSE)` is not implemented.** Only the values
  computed at fitting time are available.
- **Whole days only.** A day is either completely covered or not modeled;
  there is no partial-day fitting. See
  [Why a day may be missing from your results](#missing-days).



