---
title: "Two-Station Models: Estimating Travel Time"
author: "Two-Station Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-Station Models: Estimating Travel Time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=80)
```

```{r libs, warning=FALSE, message=FALSE}
library(streamMetabolizer)
library(dplyr)
```

This vignette continues from [Two-Station Models](two_station.html). It 
describes how to estimate the `travel.time` input, which is a required 
input for the two-station model. Travel time is not computed by 
`streamMetabolizer` — you supply it, as the `travel.time` column, in units 
of fractional days. It's specific to two-station methods: because the model tracks a parcel of
water as it moves between stations, and variable-flow reaches like
hydropeaking tailwaters have travel times that themselves change through
the day, getting this input right matters for the estimates that follow.

The variable-flow two-station (VFTS) method of Bishop et al. (2026)
compares four ways of estimating travel time, ordered here from most
data-intensive/accurate to simplest/coarsest.

## Detailed hydraulic modeling

The most accurate approach Bishop et al. compared, an agent-based
particle-tracker model producing per-timestep travel times (their
`VFTS-PT` scenario, and the basis of the more mechanistically detailed
coupled metabolic-hydraulic model of Payn et al. 2017), is also the most
computationally demanding and requires considerably more technical setup
than the other three. It's outside what `streamMetabolizer` itself
provides; see Payn et al. (2017) for an example with this level of
detail.

## Natural tracer (specific conductivity)

Rather than injecting a tracer, Bishop et al. (2026) take advantage of
conductivity variation already present in the source water. In their
site, the Colorado River below Glen Canyon Dam, conductivity varies
because the upstream reservoir (Lake Powell) is stratified: the
epilimnion and hypolimnion differ by roughly 250 µS/cm on average, with
additional fresher pulses from snowmelt inflow. A distinctive
conductivity peak recorded at the upstream sensor can be matched to the
same feature arriving hours later at the downstream sensor; the time
between them is the travel time for that water parcel.

This works when the source water has a naturally variable signal to
track, which isn't guaranteed at every site, the paper notes specific
conductivity isn't a truly conservative tracer, though it performed well
for their reach. An injected conservative tracer (e.g., bromide) is an
alternative where no natural signal exists.

A simplified demonstration, loosely following the peak-matching shown in
Bishop et al.'s Figure 2A:

```{r ts_conductivity_demo}
set.seed(2011)

# 24 hours at 15-minute resolution (96 obs/day, matching Bishop et al. 2026)
times <- seq(as.POSIXct("2011-08-02 00:00:00", tz="UTC"),
             as.POSIXct("2011-08-03 00:00:00", tz="UTC"),
             by="15 min")

# Simulated upriver specific conductivity (uS/cm), with one distinct peak
cond_up <- 650 + 20 * exp(-((as.numeric(times) - as.numeric(times[40]))^2) /
                             (2 * (3600 * 2)^2)) +
  rnorm(length(times), sd = 1.5)

# Downriver trace: same peak, arriving ~6 hours later
travel_time_true_hr <- 6
shift_steps <- round(travel_time_true_hr * 60 / 15)
cond_down <- c(rep(645, shift_steps), cond_up[1:(length(cond_up) - shift_steps)]) +
  rnorm(length(times), sd = 1.5)

# Locate the peak in each trace and compute the offset
peak_up_idx <- which.max(cond_up)
peak_down_idx <- which.max(cond_down)

travel_time_hr <- as.numeric(difftime(times[peak_down_idx], times[peak_up_idx],
                                       units = "hours"))
travel_time_hr
```

This toy example uses `which.max()` to find the peak in each trace.
Bishop et al.'s actual approach involves more careful identification of
*unambiguous* peaks (real records have multiple candidate peaks, noise,
and occasional ambiguity between upriver and downriver features), and
they only used peaks that could be confidently matched.

<!-- TODO: expand this demo with a more realistic peak-detection approach,
and/or real paired conductivity data, once available. -->

## Discharge-based prediction

Natural-tracer travel times are typically only available for a subset of
days with clearly identifiable peaks, not continuously. Bishop et al. (2026) fit
a log-linear regression between their tracer-derived travel times and
discharge, with an additional term for the rate of change in discharge
(to account for hysteresis — travel time responds differently to rising
vs. falling flow), then used that fitted relationship to predict travel
time continuously for every timestep in the record:

```
ln(travel_time) = β0 + β1 * ln(discharge) + β2 * (dQ/dt)
```

Fit to n = 1,476 tracer-derived travel times spanning discharge from
118–716 m^3/s, their best-fit relationship (at a mean dQ/dt) was:

```
ln(travel time) = 4.74 + ln(discharge) - 0.02
```

This is the `VFTS-CQ` scenario in Bishop et al. (2026), and it's the
approach that produces time-varying `travel.time` values reflecting
changing velocity with discharge. A similar site-specific model could be
developed in other sytems, but we do not recommend applying this exact fitting
outside of this example.

<!-- TODO: work through fitting an equivalent regression against a subset
of two_station_example, once paired conductivity data is available to
demonstrate with. -->

## Uniform travel time

The coarsest option: a single travel time, calculated as mean reach
length divided by mean reach velocity, applied uniformly to every day and
timestep. This is the `VFTS-MT` scenario in Bishop et al. (2026) — the
least accurate of the four approaches they compared, but the only one
requiring no tracer data at all, just basic reach dimensions and average
discharge. Bishop et al. found this approach's accuracy highly sensitive
to which single value gets chosen; supplemental runs using the 10th and
90th percentile of discharge over their study period produced
substantially different metabolism estimates than the mean-discharge run.


## The travel-time ceiling

<!-- TODO: keep this section in sync if the ceiling default or cap changes
(mm_max_travel_time_default / mm_max_travel_time_cap). -->

Travel time is subject to a ceiling, `specs$max_travel_time_hours`, which
defaults to **10 hours** and may be set as high as 12. Any day whose longest
travel time exceeds the ceiling is dropped, with a message; the remaining
days are fit normally.

The reason for dropping days with high travel times is the treatment of the
`light` column. Beyond the ceiling, a day's upstream parcel
originates before that day's own 06:00 start, and the within-day light
proportion it experienced no longer has a well-defined day total to be a
proportion of. A travel time far above the ceiling usually means the column
was supplied in the wrong units — fractional days are expected, not hours or minutes.

This same 10-hour threshold appears in Bishop et al. (2026): 43 of their
study days had discharge-based travel time estimates exceeding 10 hours
for at least one timestep, and those days were dropped from their
analysis for the same reason, light estimates would otherwise include
incident light carried over from the previous day.

## Travel-time uncertainty

<!-- TODO: expand with additional guidance. -->

Uncertainty in travel time is not propagated into the metabolism estimates.
If travel time is poorly constrained at your site, fit the model separately
with low and high travel-time estimates (as in the `VFTS-MTlow` and
`VFTS-MThi` sensitivity scenarios of Bishop et al. 2026) to bracket the
effect on GPP, ER, and K600.

### References

Bishop, I.W., Deemer, B.R., Kennedy, T.A., Payn, R.A., Hall Jr, R.O.
and Yackulic, C.B., 2026. A simplified two-station approach for modeling
metabolism in dam tailwaters subject to diel flow variation. *Limnology
and Oceanography: Methods*, p.e70066.
https://doi.org/10.1002/lom3.70066

Payn, R.A., Hall, R.O., Kennedy, T.A., Poole, G.C. and Marshall, L.A.,
2017. A coupled metabolic-hydraulic model and calibration scheme for
estimating whole-river metabolism during dynamic flow conditions.
*Limnology and Oceanography: Methods*, 15(10), pp.847-866.
https://doi.org/10.1002/lom3.10204
