---
title: "Data Preparation"
author: "Alison Appling and Bob Hall"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Requirements

A properly formatted input dataset for streamMetabolizer models has:

* exactly the right data columns and column names. Call `metab_inputs()` to see the requirements for a specific model type.
* data in the right units. See `?mm_data` for definitions of each column.

An input dataset may optionally include:

* partial days; partial days will be automatically excluded, so you don't need to do this yourself.
* non-continuous days; no current streamMetabolizer models require continuous days.

# Example

An example of a properly formatted input dataset is available in the
streamMetabolizer package - data are from French Creek in Laramie, WY, courtesy
of Bob Hall.
```{r data, message=FALSE}
library(streamMetabolizer)
dat <- data_metab(num_days='3', res='15')
```

Inspect the dimensions and column names of the data.
```{r data_check}
dim(dat)
dat[c(1,48,96,240,288),] # some example rows
```

You can get additional information about the expected format of the data in the `?metab` help document.
When preparing your own data, make sure the class and units of your data match those specified in that document.

# Exploring input data

You can use other common R packages to graphically inspect the input data. Look for outliers and oddities to ensure the quality of your data.

```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r viz_inputs_DO, fig.width=7, fig.height=3}
dat %>% 
  mutate(DO.pctsat = 100 * (DO.obs / DO.sat)) %>%
  select(solar.time, starts_with('DO')) %>%
  gather(type, DO.value, starts_with('DO')) %>%
  mutate(units=ifelse(type == 'DO.pctsat', 'DO\n(% sat)', 'DO\n(mg/L)')) %>%
  ggplot(aes(x=solar.time, y=DO.value, color=type)) + geom_line() + 
  facet_grid(units ~ ., scale='free_y') + theme_bw() +
  scale_color_discrete('variable')
```

```{r viz_inputs_other, fig.width=7, fig.height=4}
labels <- c(depth='depth\n(m)', temp.water='water temp\n(deg C)', light='PAR\n(umol m^-2 s^-1)')
dat %>% 
  select(solar.time, depth, temp.water, light) %>%
  gather(type, value, depth, temp.water, light) %>%
  mutate(
    type=ordered(type, levels=c('depth','temp.water','light')),
    units=ordered(labels[type], unname(labels))) %>%
  ggplot(aes(x=solar.time, y=value, color=type)) + geom_line() + 
  facet_grid(units ~ ., scale='free_y') + theme_bw() +
  scale_color_discrete('variable')
```

# Check the input data format

Your data need to have specific column names and units. To see what is required,
use the `metab_inputs` function to get a description of the required inputs for
a given model type. The output of metab_inputs is a table describing the
required column names, the classes and units of the values in each column, and
whether that column is required or optional. The inputs are identical for the
model types 'mle', 'bayes', and 'night', so here we'll just print the
requriements for 'mle'.
```{r data_needs}
metab_inputs('mle', 'data')
```

Also read through the help pages at `?metab` and `?mm_data` for more detailed variable definitions and requirements.

# Prepare the timestamps

To prepare your timestamps for metabolism modeling, you need to convert from the
initial number or text format into POSIXct with the correct timezone (tz), then
to solar mean time.

## Step 1: POSIXct

Convert your logger-format data to POSIXct in a local timezone (with or without
daylight savings, as long as you have that timezone scheme specified). Here are
a few examples of specific scenarios and solutions.

### Starting with numeric datetimes, e.g., from PMEs

If you have datetimes stored in seconds since 1/1/1970 at Greenwich (i.e., in UTC):
```{r}
num.time <- 1471867200
(posix.time.localtz <- as.POSIXct(num.time, origin='1970-01-01', tz='UTC'))
```

If you have datetimes stored in seconds since 1/1/1970 at Laramie, WY (i.e., in MST, no daylight savings):
```{r}
num.time <- 1471867200
(posix.time.nominalUTC <- as.POSIXct(num.time, origin='1970-01-01', tz='UTC')) # the numbers get treated as UTC no matter what tz you request
(posix.time.localtz <- lubridate::force_tz(posix.time.nominalUTC, 'Etc/GMT+7')) # +7 = mountain standard time
```


### Starting with text timestamps

If you have datetimes stored as text timestamps in UTC, you can bypass the conversion to local time and just start with UTC. Then rather than using `calc_solar_time()` in Step 2, you'll use `convert_UTC_to_solartime()`.
```{r}
text.time <- '2016-08-22 12:00:00'
(posix.time.utc <- as.POSIXct(text.time, tz='UTC'))
```

If you have datetimes stored as text timestamps in EST/EDT (with daylight savings):
```{r}
text.time <- '2016-08-22 12:00:00'
(posix.time.localtz <- as.POSIXct(text.time, format="%Y-%m-%d %H:%M:%S", tz='America/New_York'))
```

If you have datetimes stored as text timestamps in EST (no daylight savings):
```{r}
text.time <- '2016-08-22 12:00:00'
(posix.time.localtz <- as.POSIXct(text.time, format="%Y-%m-%d %H:%M:%S", tz='Etc/GMT+5'))
```

See https://en.wikipedia.org/wiki/List_of_tz_database_time_zones for a list of timezone names.


### Starting with `chron` datetimes

If you have datetimes stored in the `chron` time format in EST (no daylight savings):
```{r}
chron.time <- chron::chron('08/22/16', '12:00:00')
time.format <- "%Y-%m-%d %H:%M:%S"
text.time <- format(chron.time, time.format) # direct as.POSIXct time works poorly
(posix.time.localtz <- as.POSIXct(text.time, format=time.format, tz='Etc/GMT+5'))
```

## Step 2: Solar time

Now convert from local time to solar time. In `streamMetabolizer` vocabulary, 
`solar.time` specifically means mean solar time, the kind where every day is 
exactly 24 hours, in contrast to apparent solar time. You're ready for this step
when you have the correct time in a local timezone and `lubridate::tz(yourtime)`
reflects the correct timezone.

```{r}
lubridate::tz(posix.time.localtz) # yep, we want and have the code for EST
(posix.time.solar <- streamMetabolizer::calc_solar_time(posix.time.localtz, longitude=-106.3))
```

# Other data preparation

streamMetabolizer offers many functions to help you prepare your data for modeling. We'll add more to this section later. For now, we recommend that you explore the help pages for the following functions:

* `calc_depth`
* `calc_DO_sat`
* `calc_light`
* `convert_date_to_doyhr`
* `convert_localtime_to_UTC`
* `convert_UTC_to_solartime`
* `convert_k600_to_kGAS`
* `convert_PAR_to_SW`
