---
title: "Data Preparation"
author: "Alison Appling and Bob Hall"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{get_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Time

To prepare your timestamps for metabolism modeling, you need to convert from the
initial number or text format into POSIXct with the correct timezone (tz), then
convert to UTC, then convert to solar mean time.

## Step 1: POSIXct

Convert your logger-format data to POSIXct in a local timezone (with or without
daylight savings). Here are a few examples of specific scenarios and solutions.

### Starting with numeric datetimes, e.g., from PMEs

If you have datetimes stored in seconds since 1/1/1970 at Greenwich (i.e., in UTC):
```{r}
num.time <- 1471867200
(posix.time.localtz <- as.POSIXct(num.time, origin='1970-01-01', tz='UTC'))
```

If you have datetimes stored in seconds since 1/1/1970 at Laramie, WY (i.e., in MST, no daylight savings):
```{r}
num.time <- 1471867200
(posix.time.nominalUTC <- as.POSIXct(num.time, origin='1970-01-01', tz='UTC')) # the numbers get treated as UTC no matter what tz you request
(posix.time.localtz <- lubridate::force_tz(posix.time.nominalUTC, 'Etc/GMT+7')) # +7 = mountain standard time
```


### Starting with text timestamps

If you have datetimes stored as text timestamps in UTC:
```{r}
text.time <- '2016-08-22 12:00:00'
(posix.time.utc <- as.POSIXct(text.time, tz='UTC'))
```

If you have datetimes stored as text timestamps in EST/EDT (no daylight savings):
```{r}
text.time <- '2016-08-22 12:00:00'
(posix.time.localtz <- as.POSIXct(text.time, format="%Y-%m-%d %H:%M:%S", tz='America/New_York'))
```

If you have datetimes stored as text timestamps in EST (no daylight savings):
```{r}
text.time <- '2016-08-22 12:00:00'
(posix.time.localtz <- as.POSIXct(text.time, format="%Y-%m-%d %H:%M:%S", tz='Etc/GMT+5'))
```

See https://en.wikipedia.org/wiki/List_of_tz_database_time_zones for a list of timezone names.

## Step 2: Solar time

Now convert from local time to solar time. In `streamMetabolizer` vocabulary, 
`solar.time` specifically means mean solar time, the kind where every day is 
exactly 24 hours, in contrast to apparent solar time. You're ready for this step
when you have the correct time in a local timezone and `lubridate::tz(yourtime)`
reflects the correct timezone.

```{r}
lubridate::tz(posix.time.localtz) # yep, we want and have the code for EST
(posix.time.solar <- streamMetabolizer::calc_solar_time(posix.time.localtz, longitude=-106.3))
```

