---
title: "ODE Methods"
author: "Alison Appling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ODE Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  
```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

# Overview

This vignette demonstrates how you can choose and compare the ODE solution
method, which is the numerical algorithm used to translate from a given set of
daily metabolism parameters into a time series of dissolved oxygen predictions.

# Setup

Load streamMetabolizer and some helper packages.

```{r, message=FALSE}
library(streamMetabolizer)
library(dplyr)
library(ggplot2)
```

Get some data to work with: here we're requesting three days of data at 
30-minute resolution. Thanks to Bob Hall for the test data.

```{r}
dat <- data_metab('3','30')
```

# Numerical integration

Inspired by Song et al. 2016, we can now do several types of numerical 
integration and compare them. `lsoda` often fails to converge, but `rk4` and 
`trapezoid` perform well and very similarly to one another (and `trapezoid` is 
faster). `trapezoid` is available for both MLE and Bayesian models.

Here we fit MLE models using four different ODE methods.
```{r}
mm_euler <- metab(specs(mm_name('mle', ode_method='euler')), dat)
```

```{r}
mm_trapezoid <- metab(specs(mm_name('mle', ode_method='trapezoid')), dat)
```

```{r}
mm_rk4 <- metab(specs(mm_name('mle', ode_method='rk4')), dat) 
```

```{r}
mm_lsoda <- metab(specs(mm_name('mle', ode_method='lsoda')), dat) 
```

Now we create a data.frame to compare the above options.
```{r}
DO.standard <- rep(predict_DO(mm_rk4)$'DO.mod', times=4)
ode_preds <- bind_rows(
  mutate(predict_DO(mm_euler), method='euler'),
  mutate(predict_DO(mm_trapezoid), method='trapezoid'),
  mutate(predict_DO(mm_rk4), method='rk4'),
  mutate(predict_DO(mm_lsoda), method='lsoda')) %>%
  mutate(DO.mod.diffeuler = DO.mod - DO.standard)
```

We can plot the predictions from each method.
```{r, fig.width=6}
ggplot(ode_preds, aes(x=solar.time)) +
  geom_point(aes(y=DO.obs), color='grey', alpha=0.3) +
  geom_line(aes(y=DO.mod, color=method), size=1) +
  theme_bw()
```

To inspect the details, we can also plot the predictions as deviations from the rk4 method.
```{r, fig.width=6}
ggplot(ode_preds, aes(x=solar.time)) +
  geom_point(aes(y=pmax(-0.2, pmin(0.2, DO.mod.diffeuler)), color=method), size=1, alpha=0.8) +
  scale_y_continuous(limits=c(-0.2,0.2)) +
  ylab("Deviations from rk4 (capped at +/- 0.2)") +
  theme_bw()
```
