---
title: "Simulations"
author: "Alison Appling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{get_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=100)
```

# Overview

This vignette shows how to simulate dissolved oxygen 'observations' for the
purpose of exploring and testing metabolism models.

# Setup

Load streamMetabolizer.

```{r, messages=TRUE, warnings=TRUE, errors=TRUE}
library(streamMetabolizer)
```

Load dplyr.

```{r}
suppressPackageStartupMessages(library(dplyr))
```

Get some data to work with: here we're requesting three days of data at
15-minute resolution. Thanks to Bob Hall for the test data.

```{r}
dat <- data_metab('3', '15')
```

# Creating a Sim Model

To create a simulation model, you should 

1. Choose a model structure
2. Choose daily metabolism parameters
3. Choose the other model specifications
4. Create the model
5. Generate predictions (simulations) from the model

## 1. Choosing the model structure

You can simulate data using any of the GPP and ER functions available to MLE 
models. Simulations are done by models of type `'sim'` but otherwise take very 
similar arguments to those of an MLE model. Here we'll use a model where ER is a
function of temperature.

```{r}
name_sim_q10 <- mm_name('sim', ER_fun='q10temp')
```

## 2. Choosing the daily parameters

To simulate data, you need to specify the daily parameters beforehand. The model
structure determines which parameters are needed. There are three good ways to
learn which daily parameters you need to specify.

### A. Trial and error

To learn about parameter needs by trial and error, simply create the model with 
the equations you want but without daily inputs, ask for the parameters, and 
read the error message to get a list of parameters. It's fine to use the
defaults for the specifications for now.

```{r, error=TRUE}
mm_sim_q10_trial <- metab(specs(name_sim_q10), dat)
get_params(mm_sim_q10_trial)
```

Great: we need `GPP.daily`, `ER20`, and `K600.daily`. Now we can pick values for
those parameters and put them in a data.frame.

```{r}
params_sim_q10a <- data.frame(date=as.Date(paste0('2012-09-',18:20)), GPP.daily=2.1, ER20=-5:-3, K600.daily=16)
params_sim_q10a
```

### B. Generate parameters from another model

You can also use fitted parameters from another model as your input for a
simulation model. This method could be useful for identifying realistic
parameters and/or exploring why a model fitting process didn't work so well.

First fit an MLE model to the same data using the `GPP_fun` and `ER_fun` you 
want. It's fine (again) to use the defaults for the specifications.

```{r}
mm_mle_q10 <- metab(specs(mm_name('mle', ER_fun='q10temp')), data=dat)
```

Then ask for the parameters in the right format (without columns for uncertainty
or messages).

```{r}
params_sim_q10b <- get_params(mm_mle_q10, uncertainty='none', messages=FALSE)
params_sim_q10b
```

### C. Look at `?mm_name`

Try it. We put lots of details in the help file. Check out the documentation for
the `GPP_fun` and `ER_fun` args in particular.

```{r, eval=FALSE}
?mm_name
```

After reading the documentation you'll create a data.frame of the same format as
in options A or B.

## 3. Choosing the specifications

After choosing parameters, the next step is to choose the rest of the 
specifications. The main difference between sim models and other models is that 
you can choose values for the probability distributions of the observation 
and/or process errors. See `?specs` for details on the distribution parameters
`err.obs.sigma`, `err.obs.phi`, `err.proc.sigma`, and `err.proc.phi`.

```{r}
specs_sim_q10 <- specs(name_sim_q10, err.obs.sigma=0, err.proc.sigma=1)
specs_sim_q10
```

## 4. Creating a model

Now you can create a simulation model much as you would an MLE or Bayesian
model. We'll make two models here, one for each of the parameter sets we created
above.

```{r}
mm_sim_q10a <- metab(specs_sim_q10, dat, data_daily=params_sim_q10a)
mm_sim_q10b <- metab(specs_sim_q10, dat, data_daily=params_sim_q10b)
```

## 5. Generating predictions

Predictions and simulations are one and the same when your model is of type `sim`.

```{r, fig.height=2}
plot_DO_preds(mm_sim_q10a, y_var='conc')
plot_DO_preds(mm_sim_q10b, y_var='conc')
```


# Simulating Errors

The main purpose of simulation models is to generate DO 'observations' with
error, i.e., noise, to see whether other models can recover the underlying
parameters despite the noise.

For this section we'll use a simulation with GPP as a saturating function of
light. We'll use method B from above to choose our daily parameters.

```{r}
specs_sim_sat <- specs(mm_name('sim', GPP_fun='satlight'), err.obs.sigma=0, err.proc.sigma=1)
params_sim_sat <- get_params(metab(specs(mm_name('mle', GPP_fun='satlight')), data=dat), uncertainty='none', messages=FALSE)
```

## Innovative errors

By default, simulations generate new noise each time you request predictions.

```{r, fig.height=2}
mm_sim_sat_i <- metab(specs_sim_sat, dat, data_daily=params_sim_sat)
plot_DO_preds(mm_sim_sat_i, y_var='conc')
plot_DO_preds(mm_sim_sat_i, y_var='conc')
```

## Fixed errors

Alternatively, you can revise the value of `sim.seed` to be a number (any
number) and then the simulation produces the same noise each time.

```{r, fig.height=2}
mm_sim_sat_f <- metab(revise(specs_sim_sat, sim.seed=47), dat, data_daily=params_sim_sat)
plot_DO_preds(mm_sim_sat_f, y_var='conc')
plot_DO_preds(mm_sim_sat_f, y_var='conc')
```


# Inspecting Models

We'll use a slightly longer dataset here to demonstrate the potential for random
noise at the levels of both the observations (every time you run `predict_DO()`)
and the daily parameters (every time you define `data_daily`).

```{r}
dat <- data_metab('10', '30')
params <- data.frame(date=as.Date(paste0('2012-09-',18:27)), Pmax=rnorm(10, 6, 2), alpha=rnorm(10, 0.01, 0.001), ER20=rnorm(10, -4, 2), K600.daily=16)
specs <- specs(mm_name('sim', GPP_fun='satlight', ER_fun='q10temp'), err.obs.sigma=0.2, err.proc.sigma=1)
mm <- metab(specs, data=dat, data_daily=params)
```

Sim models print out their parameters with asterisks to denote that the values 
are fixed rather than fitted. 

```{r}
mm
```

Sim models produce daily estimates of GPP and ER, which should help in choosing
simulation parameters. The GPP and ER predictions have no error bars because
they're direct calculations from the daily parameters.

```{r}
plot_metab_preds(mm)
```
